The imprint of global magma oceans on exoplanet demographics

### 2023-03-06 src/tex/ms.tex
discuss assumptions regarding water mass fraction

\begin{figure}
    \begin{centering}
        \includegraphics[width=\hsize]{figures/radiuscomparison.pdf}
    \end{centering}
    \begin{centering}
        \includegraphics[width=\hsize]{figures/radiusevolution.pdf}
        \caption{
            \textit{Top}: Mass-radius relationships for a water mass fraction $x_{H_2O}= \wrr$ and different planet states. Green: planets with a solid mantle and no steam atmosphere. Dashed: planets with steam atmospheres. Blue: planets with steam atmospheres and including the effect of water incorporation in the melt.
            \textit{Bottom}: Radius evolution of different planet types, illustrating degeneracies and potential for confusion among planet classes. Shown is a schematic time evolution of the transit radius normalized to the atmosphere-free radius $R_\mathrm{0}$ for different scenarios. Planets can move between planet classes through processes such as atmospheric loss and desiccation, which ultimately ends a runaway greenhouse phase on a timescale dependent on a planet's water content.}
        \label{fig:radiusevolution}
    \end{centering}
\end{figure}
Figure~\ref{fig:radiusevolution} shows the different mass-radius relations of the three cases for a fiducial water mass fraction of $x_{H_2O}= \wrr$.
Steam atmospheres cause a significant radius increase, which is slightly reduced when water incorporation in the melt is considered.
In the following, we only distinguish between the non-runaway greenhouse and wet melt scenarios.

@@

\section{Results}
\subsection{Statistical signature of the runaway greenhouse threshold}
To characterize the population-level imprint of individual radius changes, we generated a generic planet population with an injected runaway greenhouse effect assuming a water fraction $x_\mathrm{H_2O} = \wrr$.
\begin{figure}
    \begin{centering}
        \includegraphics[width=\hsize]{figures/HnullHmo.pdf}
        \caption{Synthetic planets above and below the runaway greenhouse threshold.
        \textit{Top}: Planet state as a function of orbital period. Planets with and without a runaway greenhouse climate mix and are not distinguishable in period space.
        \textit{Bottom}: Radii of synthetic planets with injected radius deviation as a function of net instellation. Only the planets marked as transiting are observable. Above the runaway greenhouse threshold $S_\mathrm{thresh} =\SI{280}{\watt\per\meter\squared} $, the radii of a fraction of planets are inflated by the amount indicated with gray lines. The sharp boundary at $S_\mathrm{thresh}$ causes a discontinuity in the average planet radius (blue). This runaway greenhouse hypothesis can be tested against the null hypothesis $H_0$ (green), where average radii are independent of instellation.}
        \label{fig:HnullHmo}
    \end{centering}
\end{figure}
Figure~\ref{fig:HnullHmo} shows the resulting planetary radii.
When ordered in orbital period space, the different planet types overlap, diluting the demographic imprint.
With net instellation as an independent variable, planets above and below the runaway greenhouse threshold separate.
We also show the predictions of observable average planet radii from the statistical hypotheses defined above.
%The value for the average radius \textit{outside} the runaway greenhouse regime was computed from our nominal observed sample, whereas the radius change in the magma ocean case compared to the null hypothesis is a prediction from the magma ocean model, evaluated on the same sample.
Within the runaway greenhouse limit, an average radius change of \avgRadiusChange\ occurs.

\subsection{Testability of the runaway greenhouse hypothesis}\label{sec:res_testability}
\begin{figure}[ht!]
%    \script{figurescript.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/optimistic_R-S.pdf}
        \caption{
        Detection of the runaway greenhouse threshold in the optimistic case.
        From simulated radius and instellation measurements of a large ($N = \var{N_optimistic}$) survey, we compute the moving average (blue confidence intervals) and fit the runaway greenhouse hypothesis to it (Eqn.~\ref{eq:rgh_hypo}, random draws from the posterior in red).
            The pattern is detected with high significance.
        }
        \label{fig:optimistic_R-S}
    \end{centering}
\end{figure}
Figure~\ref{fig:optimistic_R-S} shows a prototypical statistical detection of the runaway greenhouse transition.
For the sake of clarity, we chose an optimistic scenario where the sample is large ($N = \var{N_optimistic}$) and the measurement uncertainties are small ($\sigma_{R} = \sigmaR, \sigma_{S} = \sigmaS$).
We assumed that the fraction of those planets irradiated stronger than $S_\mathrm{thresh}$ that have runaway greenhouse climates is \frgh, and we chose a water mass fraction of \wrr\ for each planet.
In this case, the pattern was detected with high significance ($\Delta \ln \mathcal{Z} \approx 100$).

@@

\subsubsection{Constraining the threshold instellation}
\begin{figure}[ht!]
%    \script{figurescript.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/S_thresh_posteriors.pdf}
        \caption{
            Retrieved posterior distributions of the threshold instellation for different survey realizations.
            All cases assume $x_{H_2O} = \wrr$ and a planet sample size $N=100\pm10$; The fraction of planets with runaway greenhouse climates varies across rows.
            Orange lines show the true value of the injected signal.
            Accuracy and precision of the constraint on $S_\mathrm{thresh}$ generally improve with higher $f_\mathrm{rgh}$.
            Bulk density-based inferences improve the constraints, and M~dwarf samples yield the highest accuracy.
        }
        \label{fig:posterior_surveys}
    \end{centering}
\end{figure}
A key constraint resulting from a successful detection is a measurement of $S_{thresh}$.
Here, we assess the ability of different mission concepts explored above to constrain this parameter.
Figure~\ref{fig:posterior_surveys} shows posterior distributions from a grid of inferences together with the true values of the injected signal for different fractions $f_\mathrm{rgh}$.
We consider three cases: only radius measurements, radius and mass measurements, and radius and mass measurements of only planets orbiting M~dwarfs.
The simulations otherwise represent the simulated PLATO example described above.
We chose a planetary sample size of $N=100\pm10$, as this was found to be a threshold case in Sect.~\ref{sec:res_samplesize}.

@@

>>>>

\begin{figure}
    \begin{centering}
        \includegraphics[width=\hsize]{figures/radiuscomparison.pdf}
    \end{centering}
    \begin{centering}
        \includegraphics[width=\hsize]{figures/radiusevolution.pdf}
        \caption{
            \textit{Top}: Mass-radius relationships for a water mass fraction $x_{H_2O}= \var{wrr}$ and different planet states. Green: planets with a solid mantle and no steam atmosphere. Dashed: planets with steam atmospheres. Blue: planets with steam atmospheres and including the effect of water incorporation in the melt.
            \textit{Bottom}: Radius evolution of different planet types, illustrating degeneracies and potential for confusion among planet classes. Shown is a schematic time evolution of the transit radius normalized to the atmosphere-free radius $R_\mathrm{0}$ for different scenarios. Planets can move between planet classes through processes such as atmospheric loss and desiccation, which ultimately ends a runaway greenhouse phase on a timescale dependent on a planet's water content.}
        \label{fig:radiusevolution}
    \end{centering}
\end{figure}
Figure~\ref{fig:radiusevolution} shows the different mass-radius relations of the three cases for a fiducial water mass fraction of $x_{H_2O}= \var{wrr}$.
Steam atmospheres cause a significant radius increase, which is slightly reduced when water incorporation in the melt is considered.
In the following, we only distinguish between the non-runaway greenhouse and wet melt scenarios.

@@

\section{Results}
\subsection{Statistical signature of the runaway greenhouse threshold}
To characterize the population-level imprint of individual radius changes, we generated a generic planet population with an injected runaway greenhouse effect assuming a water fraction $x_\mathrm{H_2O} = \var{wrr}$.
\begin{figure}
    \begin{centering}
        \includegraphics[width=\hsize]{figures/HnullHmo.pdf}
        \caption{Synthetic planets above and below the runaway greenhouse threshold.
        \textit{Top}: Planet state as a function of orbital period. Planets with and without a runaway greenhouse climate mix and are not distinguishable in period space.
        \textit{Bottom}: Radii of synthetic planets with injected radius deviation as a function of net instellation. Only the planets marked as transiting are observable. Above the runaway greenhouse threshold $S_\mathrm{thresh} =\SI{280}{\watt\per\meter\squared} $, the radii of a fraction of planets are inflated by the amount indicated with gray lines. The sharp boundary at $S_\mathrm{thresh}$ causes a discontinuity in the average planet radius (blue). This runaway greenhouse hypothesis can be tested against the null hypothesis $H_0$ (green), where average radii are independent of instellation.}
        \label{fig:HnullHmo}
    \end{centering}
\end{figure}
Figure~\ref{fig:HnullHmo} shows the resulting planetary radii.
When ordered in orbital period space, the different planet types overlap, diluting the demographic imprint.
With net instellation as an independent variable, planets above and below the runaway greenhouse threshold separate.
We also show the predictions of observable average planet radii from the statistical hypotheses defined above.
%The value for the average radius \textit{outside} the runaway greenhouse regime was computed from our nominal observed sample, whereas the radius change in the magma ocean case compared to the null hypothesis is a prediction from the magma ocean model, evaluated on the same sample.
Within the runaway greenhouse limit, an average radius change of \avgRadiusChange\ occurs.

\subsection{Testability of the runaway greenhouse hypothesis}\label{sec:res_testability}
\begin{figure}[ht!]
%    \script{figurescript.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/optimistic_R-S.pdf}
        \caption{
        Detection of the runaway greenhouse threshold in the optimistic case.
        From simulated radius and instellation measurements of a large ($N = \var{N_optimistic}$) survey, we compute the moving average (blue confidence intervals) and fit the runaway greenhouse hypothesis to it (Eqn.~\ref{eq:rgh_hypo}, random draws from the posterior in red).
            The pattern is detected with high significance.
        }
        \label{fig:optimistic_R-S}
    \end{centering}
\end{figure}
Figure~\ref{fig:optimistic_R-S} shows a prototypical statistical detection of the runaway greenhouse transition.
For the sake of clarity, we chose an optimistic scenario where the sample is large ($N = \var{N_optimistic}$) and the measurement uncertainties are small ($\sigma_{R} = \sigmaR, \sigma_{S} = \sigmaS$).
We assumed that the fraction of those planets irradiated stronger than $S_\mathrm{thresh}$ that have runaway greenhouse climates is \frgh, and we chose a water mass fraction of \var{wrr} for each planet.
In this case, the pattern was detected with high significance ($\Delta \ln \mathcal{Z} \approx 100$).

@@

\subsubsection{Constraining the threshold instellation}
\begin{figure}[ht!]
%    \script{figurescript.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/S_thresh_posteriors.pdf}
        \caption{
            Retrieved posterior distributions of the threshold instellation for different survey realizations.
            All cases assume $x_{H_2O} = \var{wrr}$ and a planet sample size $N=100\pm10$; The fraction of planets with runaway greenhouse climates varies across rows.
            Orange lines show the true value of the injected signal.
            Accuracy and precision of the constraint on $S_\mathrm{thresh}$ generally improve with higher $f_\mathrm{rgh}$.
            Bulk density-based inferences improve the constraints, and M~dwarf samples yield the highest accuracy.
        }
        \label{fig:posterior_surveys}
    \end{centering}
\end{figure}
A key constraint resulting from a successful detection is a measurement of $S_{thresh}$.
Here, we assess the ability of different mission concepts explored above to constrain this parameter.
Figure~\ref{fig:posterior_surveys} shows posterior distributions from a grid of inferences together with the true values of the injected signal for different fractions $f_\mathrm{rgh}$.
We consider three cases: only radius measurements, radius and mass measurements, and radius and mass measurements of only planets orbiting M~dwarfs.
The simulations otherwise represent the simulated PLATO example described above.
We chose a planetary sample size of $N=100\pm10$, as this was found to be a threshold case in Sect.~\ref{sec:res_samplesize}.

@@

\subsubsection{Water mass fraction}
The predominant mass fractions of water, which sensitively controls the atmospheric state of a rocky exoplanet, is poorly constrained.
Inferred water contents in the literature range from upper limits on the order $10^{-5}$ to ``water worlds'' with tens of percent weight fraction~\citep[e.g.,][]{Rogers2010,Unterborn2018,Mousis2020,Agol2021,Luque2022}, all of which are within the realm of theoretical predictions~\citep{Mulders2015b,Sato2016,Jin2018,Bitsch2019b,Venturini2020,Emsenhuber2021b,Schlecker2021}.
Our nominal case assumes a water mass fraction of $x_\mathrm{H_2O} = \var{wrr}$.
This can be considered a conservative choice that is unlikely to introduce a systematic overestimation of the statistical radius inflation trend, also since the hypothesis tests are only marginally sensitive on $x_\mathrm{H_2O}$ as long as it is $\gtrsim 10^{-4}$ (see Sect.~\ref{sec:res_testability}).
Cases of pure rocky composition and very low volatile contants can be considered absorbed by the dilution factor $f_\mathrm{rgh}$.
Assuming a distribution of water mass fractions instead of a fixed value would thus not significantly change our results.
